---
title: "random_forest_diabetes"
author: "Noelia Caba Martin"
date: "2026-03-17"
output: html_document
---

## Introduction

Random Forest is an ensemble machine learning algorithm that builds 
multiple decision trees and combines their predictions through majority 
voting. It is widely used in clinical research for classification and 
prediction tasks due to its high accuracy, robustness to outliers, and 
ability to handle non-linear relationships.

In this analysis, we use Random Forest to build a **diabetes prediction 
model** using clinical and demographic variables from the NHANES dataset. 
The model will be evaluated using standard classification metrics.

**Research question:** Can we accurately predict diabetes status from 
clinical and demographic variables?

The main objectives are:

- Prepare and split the data into training and test sets
- Train a Random Forest classification model
- Evaluate model performance using accuracy, sensitivity, specificity 
and AUC-ROC
- Identify the most important predictors of diabetes

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
```{r libraries}
library(randomForest)
library(caret)
library(pROC)
library(vip)
library(tidyverse)
library(ggplot2)
library(NHANES)
```

## Data Preparation

We use the same NHANES dataset from the previous project, applying 
the same cleaning steps to ensure consistency. The dataset is then 
split into **training (70%)** and **test (30%)** sets.

The training set is used to build the model, while the test set is 
used to evaluate its performance on unseen data. This prevents 
overfitting — a common problem where a model performs well on 
training data but poorly on new data.

```{r data-preparation}
# Load and clean data
set.seed(123)

data(NHANES)

nhanes_rf <- NHANES %>%
  select(Gender, Age, Race1, BMI, BPSysAve, 
         BPDiaAve, TotChol, PhysActive, 
         SleepHrsNight, Diabetes) %>%
  distinct() %>%
  filter(!is.na(Diabetes)) %>%
  filter(!is.na(BMI)) %>%
  filter(!is.na(BPSysAve)) %>%
  filter(!is.na(TotChol)) %>%
  filter(!is.na(PhysActive)) %>%
  drop_na()

# Check dimensions and class balance
dim(nhanes_rf)
table(nhanes_rf$Diabetes)

# Split into training and test sets (70/30)
train_index <- createDataPartition(nhanes_rf$Diabetes, 
                                   p = 0.7, 
                                   list = FALSE)
train_data <- nhanes_rf[train_index, ]
test_data <- nhanes_rf[-train_index, ]

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
```

The cleaned dataset contains **4,515 participants** with complete data 
for all selected variables. The dataset shows class imbalance, with 
89.5% non-diabetic and 10.5% diabetic participants.

The data was split into:
- **Training set:** 3,162 observations (70%)
- **Test set:** 1,353 observations (30%)

`createDataPartition()` ensures that the class distribution is 
preserved in both sets (stratified splitting).

## Model Training

### Class Imbalance

Before training, it is important to address the class imbalance. 
If left unaddressed, the model may achieve high overall accuracy 
simply by predicting "No diabetes" for all participants, without 
learning meaningful patterns.

We will use **class weights** to penalize misclassification of the 
minority class (diabetes) more heavily than the majority class.

```{r train-model}
# Define class weights to handle imbalance
class_weights <- c("No" = 1, "Yes" = 10)

# Train Random Forest model
set.seed(123)
rf_model <- randomForest(
  Diabetes ~ .,
  data = train_data,
  ntree = 500,
  mtry = 3,
  importance = TRUE,
  classwt = class_weights
)

# Print model summary
print(rf_model)
```

The Random Forest model was trained on 3,162 observations using 500 
trees and 3 variables per split.

The OOB error rate of **10.5%** provides an unbiased estimate of 
model performance. However, the confusion matrix reveals a common 
challenge with imbalanced datasets: while the model correctly 
classifies 99.4% of non-diabetic cases, it only identifies 5.7% 
of diabetic cases correctly.

This highlights the importance of using multiple evaluation metrics 
beyond overall accuracy when dealing with imbalanced classification problems.

## Model Evaluation

We evaluate the model on the **test set** — data the model has never 
seen during training. This gives an unbiased estimate of real-world 
performance.

We use the following metrics:

- **Accuracy:** overall proportion of correct predictions
- **Sensitivity (Recall):** proportion of diabetic cases correctly identified
- **Specificity:** proportion of non-diabetic cases correctly identified
- **AUC-ROC:** Area Under the ROC Curve — measures the model's ability 
to discriminate between classes regardless of threshold. Ranges from 
0.5 (random) to 1.0 (perfect)

```{r model-evaluation}
# Make predictions on test set
predictions <- predict(rf_model, test_data)
probabilities <- predict(rf_model, test_data, type = "prob")

# Confusion matrix
conf_matrix <- confusionMatrix(predictions, test_data$Diabetes, 
                                positive = "Yes")
print(conf_matrix)

# ROC curve
roc_obj <- roc(test_data$Diabetes, probabilities[, "Yes"])

# Plot ROC curve
plot(roc_obj,
     main = paste("ROC Curve - AUC:", round(auc(roc_obj), 3)),
     col = "#E64B35",
     lwd = 2,
     print.auc = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey")
```

The model evaluation on the test set reveals:

- **AUC-ROC = 0.779:** the model shows good discriminative ability, 
substantially better than random classification (AUC = 0.5)
- **Balanced Accuracy = 0.537:** reflects the challenge of the 
imbalanced dataset

The ROC curve shows that the model can achieve reasonable sensitivity 
while maintaining acceptable specificity, making it potentially useful 
as a screening tool for diabetes risk.

## Variable Importance

One of the key advantages of Random Forest over other machine learning 
algorithms is its ability to quantify the **importance of each variable** 
in making predictions.

Variable importance is measured by:

- **Mean Decrease Accuracy:** how much the model accuracy decreases 
when a variable is randomly permuted (shuffled). Higher values = 
more important variable
- **Mean Decrease Gini:** how much each variable contributes to the 
purity of the nodes in the trees

```{r variable-importance}
# Plot variable importance
vip(rf_model,
    num_features = 10,
    aesthetics = list(fill = "#2E9FDF", color = "#2E9FDF", alpha = 0.8)) +
  theme_minimal() +
  labs(title = "Variable Importance - Random Forest",
       subtitle = "Top 10 predictors of diabetes",
       x = "Importance", y = "Variable")
```

The variable importance plot reveals that **Age** is by far the strongest 
predictor of diabetes in this model, followed by Total Cholesterol, 
Race, and BMI.

Notably, Total Cholesterol ranked as the second most important variable 
despite not showing a statistically significant difference between groups 
in the univariate analysis (Table 1). This highlights a key advantage 
of Random Forest: its ability to capture **non-linear relationships and 
interactions between variables** that traditional statistical tests may miss.

Physical activity, diastolic blood pressure, and sleep hours show 
relatively low importance, suggesting they contribute little additional 
predictive information beyond age and metabolic variables.

## Conclusions

This analysis demonstrates the application of Random Forest machine 
learning to predict diabetes status from clinical and demographic 
variables in the NHANES dataset.

**Key findings:**

1. The Random Forest model achieved an **AUC-ROC of 0.779**, indicating 
good discriminative ability between diabetic and non-diabetic participants 
using only basic clinical variables.

2. **Age** was the strongest predictor of diabetes, consistent with the 
well-established increase in type 2 diabetes prevalence with age.

3. **Total Cholesterol** ranked as the second most important predictor 
despite not showing significant differences in univariate analysis, 
demonstrating the ability of Random Forest to detect complex non-linear 
relationships.

4. **BMI and Race** were also important predictors, consistent with 
known risk factors for type 2 diabetes.

5. Class imbalance (89.5% non-diabetic vs 10.5% diabetic) was addressed 
using class weights, though it remained a challenge for detecting 
diabetic cases.

**Limitations:**

- Class imbalance limits the model's sensitivity for detecting diabetic 
cases — more advanced resampling techniques (SMOTE) could improve this
- The model was trained on cross-sectional data, preventing causal 
inference
- External validation on an independent dataset would be needed before 
clinical application
- Only 10 variables were included — additional clinical variables 
(HbA1c, fasting glucose) would likely improve performance

**Future directions:**

- Apply SMOTE (Synthetic Minority Oversampling Technique) to address 
class imbalance
- Compare Random Forest with other algorithms (XGBoost, Logistic 
Regression)
- Perform hyperparameter tuning using cross-validation
- Validate the model on an independent clinical dataset